---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Here are the packages I am using for this assignment:

```{r, echo=TRUE, warning=FALSE}
library(tidyverse)
library(lubridate)
```

#### 1 & 2. Here is my code to load and process the data: 

```{r}


setwd("C:/Users/207014104/Desktop/DataScience/Reproduceable Research/Project1/repdata_data_activity")

fileName <- "activity.csv"

#read the csv once it has been unzipped in the working directory
activity <- as_tibble(read.csv(fileName))
```


## What is mean total number of steps taken per day?

#### 1. Calculate the total number of steps taken per day

```{r}
dailySteps <- group_by(activity, date) %>%
        summarize(totalSteps = sum(steps))
print(dailySteps)
```

#### 2. Make a histogram of the total number of steps taken each day

```{r}
ggplot(dailySteps, aes(x = totalSteps)) + 
        geom_histogram(binwidth = 2500, #set bins to width of 2500 steps
                       fill = "white", #make the fill of bars white
                       color = "black" #make outline of bars black
                       ) +
        labs(title = "Histogram of total steps each day (missing values omitted)", 
             x = "Count of days", 
             y = "Number of steps per day") 

```

#### 3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
meanOriginal <- mean(dailySteps$totalSteps, na.rm = TRUE)
medianOriginal <- median(dailySteps$totalSteps, na.rm = TRUE)

cat("The mean is:", meanOriginal)
cat("The median is:", medianOriginal)
```

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval and the average number of steps taken averaged across all days

```{r}
#group the data by interval and summarize by mean steps / interval
dailyProfile <- group_by(activity, interval) %>%
        summarize(intervalSteps = mean(steps, na.rm=TRUE))

# plot the 
ggplot(dailyProfile, aes(x = interval, y = intervalSteps)) + geom_line(size = 1) +
        labs(title = "Timeseries plot of average steps taken", 
             x = "5-minute interval", 
             y = "Average steps")
```

2. Which 5-minute interval contains the maximum number of steps?

```{r}
max <- dailyProfile[which.max(dailyProfile$intervalSteps),]
print(max)
cat("The interval with the max average number of steps is", max$interval)

```

## Imputing missing values

#### 1. Calculate and report the total missing values in the dataset

```{r}
#summarize the number of records in each column that are NAs
summary(is.na(activity))

#calculate the number of NA rows by subtracting the complete cases from total number of rows
missing <- nrow(activity) - sum(complete.cases(activity))

cat("Total missing values:", missing)

#what % of the total values for steps is NA
portion <- mean(is.na(activity$steps))
cat(missing, "is", portion, "of total measurements")

```

#### 2. Devise a strategy for filling in the missing values. 

My approach is to fill the mean value for the corresponding interval for each NA, using the summary data calculated above for dailyProfile

#### 3. Create a new dataset that is equal to the original but with missing data filled in

```{r}

#join the data frame with mean values for each interval to the activity data
filled <- left_join(activity, dailyProfile, by = "interval")

#subset the na values and give it the value from the mean value interval column
filled$steps[is.na(filled$steps)] <- filled$intervalSteps[is.na(filled$steps)]

#remove the mean column so that the data is the same format as original
filled <- select(filled, -intervalSteps)

print(filled)

```

#### 4. Make a histogram of the total number of steps taken each day with the filled NA data; and calculate the new mean and median

```{r}

dailySteps2 <- group_by(filled, date) %>%
        summarize(totalSteps = sum(steps))
print(dailySteps2)

ggplot(dailySteps2, aes(x = totalSteps)) + 
        geom_histogram(binwidth = 2500, #set bins to width of 2500 steps
                       fill = "white", #make the fill of bars white
                       color = "black" #make outline of bars black
                       ) +
        labs(title = "Histogram of total steps each day 
             (missing values filled w/ hourly mean)", 
             x = "Count of days", 
             y = "Number of steps per day") 

#mean & median, not removing NA's to confirm there are none
meanFilled <- mean(dailySteps2$totalSteps, na.rm = FALSE)
medianFilled <- median(dailySteps2$totalSteps, na.rm = FALSE)

cat("The new mean value is:", meanFilled)
cat("The new median value is:", medianFilled)

```
#### Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}

cat("The original mean value was:", meanOriginal)
cat("The original mean value was:", medianOriginal)
```

You can see that the mean remains the same.  
The median, however is slightly larger and now equal to the mean. 

Imputing the missing data can contribute bias to the results because measurements are now counted as more than a single reading.  Any error in the mean values will be increased by weighting it more heavily. 

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with 2 levels: "weekday" and "weekend"

```{r}
filled$date <- as_date(filled$date)

weekend <- c("Sat", "Sun")

filled <- mutate(filled, dayOfWeek = wday(date, label = TRUE)) %>%
        mutate(weekend = dayOfWeek %in% weekend)

filled$weekend <- factor(filled$weekend, 
                         levels = c(TRUE, FALSE), 
                         labels = c("weekend", "weekday")
                        )

print(filled)

```

#### 2. Make a panel plot containing a timeseries plot of the 5-min interval and the average number of steps taken for weekdays and weekends

```{r}
wkdayEnd <- group_by(filled, weekend, interval) %>%
        summarize(avg = mean(steps))

panel <- ggplot(wkdayEnd, aes(x = interval, y = avg, color=weekend)) + 
        facet_grid(weekend ~ .) + 
        geom_line(size = 1) +
        labs(title = "Timeseries plot of average steps taken on weekends and weekdays", 
             x = "5-minute interval", 
             y = "Average steps") +
        theme(legend.position="top") +
        theme(legend.title=element_blank())
        
plot(panel)

```
